Variability as a better characterization of Shannon entropy

### 2021-01-05 Articles/2019-ShannonEntropy/Submissions/EJP/EJP-ShannonEntropyIsVariability.tex
Fix bibliography formatting and add three EJP physics education references

The goal of this work is to give a comprehensive introduction to the concept of Shannon entropy, the expression $-\sum p_i \log p_i$, which is well established in such disparate fields as computer science, communication theory, physics \cite{Jaynes1957-1, Grandy} ecology and economics. The concept of entropy in physics, despite being treated extensively in the physics education literature (e.g. \cite{LeffPartI, LeffPartII, LeffPartIII, LeffPartIV, LeffPartV, carson2002undergraduate, baierlein1994entropy, geller2014entropy}), is still surrounded by a great deal of confusion \cite{Swendsen, Styer}. Partly because, at a fundamental level, there is still disagreement on what is the ``correct'' notion of entropy, particularly out of equilibrium \cite{jarzynski2017stochastic, goldstein2019gibbs, maes2003time}. Partly because the same word ``entropy'' is used to refer to concepts (e.g. topological entropy, thermodynamic entropy, graph entropy, R\'{e}nyi entropy, information entropy) that have different definitions and various degrees of overlap.  We will focus on the notion of Shannon entropy, by which we mean the use of $- \sum_i p_i \log p_i$ in the broadest characterization, which has received less attention in physics education. We will see that the Shannon entropy can be given a crisp and precise characterization and will explore its relation to the Gibbs and Boltzmann entropies in statistical mechanics.\footnote{We will not discuss its relation to thermodynamic entropy. Thermodynamic entropy is defined on the notions of heat and work while the Shannon entropy is defined on the notion of a distribution. The contexts are different and relating them goes beyond the scope of this work. } The same material is also presented in a series of videos \cite{videos}.

>>>>

The goal of this work is to give a comprehensive introduction to the concept of Shannon entropy, the expression $-\sum p_i \log p_i$, which is well established in such disparate fields as computer science, communication theory, physics \cite{Jaynes1957-1, Grandy} ecology and economics. The concept of entropy in physics, despite being treated extensively in the physics education literature (e.g. \cite{baierlein1994entropy, carson2002undergraduate, Santillan_2008, LeffPartI, LeffPartII, LeffPartIII, LeffPartIV, LeffPartV, Benguigui_2013, geller2014entropy, Tiwary_2020}), is still surrounded by a great deal of confusion \cite{Swendsen, Styer}. Partly because, at a fundamental level, there is still disagreement on what is the ``correct'' notion of entropy, particularly out of equilibrium \cite{jarzynski2017stochastic, goldstein2019gibbs, maes2003time}. Partly because the same word ``entropy'' is used to refer to concepts (e.g. topological entropy, thermodynamic entropy, graph entropy, R\'{e}nyi entropy, information entropy) that have different definitions and various degrees of overlap.  We will focus on the notion of Shannon entropy, by which we mean the use of $- \sum_i p_i \log p_i$ in the broadest characterization, which has received less attention in physics education. We will see that the Shannon entropy can be given a crisp and precise characterization and will explore its relation to the Gibbs and Boltzmann entropies in statistical mechanics.\footnote{We will not discuss its relation to thermodynamic entropy. Thermodynamic entropy is defined on the notions of heat and work while the Shannon entropy is defined on the notion of a distribution. The contexts are different and relating them goes beyond the scope of this work. } The same material is also presented in a series of videos \cite{videos}.
